Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-11-30 leadership_across_cultures_contexts.Rmd
Working in entire MS.
Added gg visreg plots in exploratory models. Enhanced group by subsistence plot labels.

Set varaible distribution plot replication to 10,000.

We first briefly review current theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership for both leaders and followers across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study, before presenting results and interpretations. Through novel empirical and exploratory analyses encompassing the full range of cultural diversity and social contexts, we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches.

@@

Leadership is a universal feature of human groups; among all societies there are particular contexts in which one individual maintains greater influence over group behavior and outcomes. Additionally, under the broad definition of leadership given above, parents can be defined as leaders of nuclear families. Strong community leadership, however, is common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @lewis_leaders_1974].

The nature of leadership across human societies and social contexts is highly diverse and the emergence of leaders and the functions they serve are closely tied to social and political structure. Among ethnographically described small-scale societies which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and tends to emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect and greater knowledge and social support to achieve or maintain community influence. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of conflict resolution and collective actions. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers. Representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders.

Within communities, leaders may emerge whenever groups of individuals form. The role of leaders in ritual contexts, such as in initiations, and in providing social services, including facilitating marriages and organizing social activities, are also expected to be common functions leaders serve.

Increased sedentarization and defensibility of resources is often associated with both social stratification and an institutionalized managerial role and stronger authority of community leaders. Among horticulturalists and agriculturalists managerial leadership can promote for more efficient use of shared resources such as water reserves and cooperative labor. Pastoralists and coastal populations often face similar pressures surrounding grazing lands and fishing access. In the wake of economic intensification, the function of leadership becomes increasing associated with the distribution of resources.

Defensibility of resources necessarily creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on dispersed, less defensible resources.

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies. Other public service and nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders CITE BASS. These positions typically are associated with institutionalized responsibilities and specific functions associated with their role. The structure of such organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At the higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization. They are not often involved in resolving immediate conflicts, such as disputes between individuals or nested bodies, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

Scholars within managerial studies have long discussed what leaders and managers do and have developed multiple models of the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [@bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group function and overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leadership functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, creating structures for accomplishing group goals, defending institutional integrity, and maintain internal consistency.

@@

@@

Despite immense variation, there do seem to be some commonalities among the qualities of community leaders. Across human societies, when strong community leaders emerge, they tend to be high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have differential influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006]. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have increased influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project including data from over 15,000 middle managers across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts. Other attributes of charismatic leadership found to be culturally contingent included ambitiousness, self-effacing, enthusiasm, and risk-taking [@den_hartog_culture_1999].

@@

Among societies with institutionalized social stratification, such as chiefdoms and agricultural communities, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004]. Managerial elites, who maintain influence over community decisions are endowed with a high degree of ideological status associated with the role and kin group [@stanish_evolution_2010]. In such contexts community leaders often lack coercive authority, but maintain strong influence through ideological systems and ritual or religious status and prowess inaccessible to non-elites [@stanish_evolution_2010]. The association of formal community leadership and supernatural abilities is likely an important precursor of institutionalized authority and social stratification.

Within state-level societies including non-industrial kingdoms leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social and political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda.

Across disciplines many theoretical models and empirical findings have implicated high intelligence as an important quality positively associated with effective leadership [@johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009]. Recent evidence from within managerial studies suggests a curvilinear effect in that exceptionally intelligence individuals at high end of the distribution may not be the most effective leaders in contemporary business settings [@antonakis_can_2017].

This theoretical development ultimately produced perspectives emphasizing that organizations become larger, increasing hierarchical, adm more complex leaders... (3078 Bass) -- INTELLIGENCE and COGNITIVE COMPLEXITY IN LARGER INSTITUIONS

@@

The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners.

Institutionalized leadership positions are also often, if not universally, associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional, but is widely associated with numerous social, material, and reproductive benefits.

Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need. Across many industries and organizations leadership positions are associated with increased pay. High social status, material wealth, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006].

@@

Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel metaanalysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

@von_rueden_leadership_2015 review a large body of theory and ethnography and outline four possible mechanisms by which leaders in small-scale societies might offset the potential costs associated with various leadership roles; leaders may, (1) facilitate collective actions that yield net benefits to themselves or their close kin; (2) claim a fee for their services, such a tax or greater share of returns; (3) receive reciprocal exchange in other currencies from followers who have benefited from their services; or (4) may gain other social or reproductive benefits by signaling their high quality as a mate and social partner.

Evolutionary models of leadership generally assume the benefits of leadership outweigh the costs, for both leaders and followers. However, the unique costs associated with various leadership positions have received less attention, relative to the benefits of leadership, across disciplines.

Group responsibilities necessarily impose opportunity costs on individual priorities and agendas, given leadership responsibilities require investments of time, energy, and resources that could be invested elsewhere.

Mediating conflicts between individuals can be physically and socially costly. Devising and proposing compromises and optimal outcomes can be cognitively challenging and the solutions leaders propose are unlikely to be equally satisfactory for everyone involved. Therefore, leaders stand to become the targets of animosity and social ridicule. In the absence of state-based systems and institutionalized inequalities, all group members are on relative equal footing and equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including overt physical aggression [@boehm_purposive_2008]. Physically formidable individuals with strong social networks are likely to be able o bear these costs more-so than their counterparts and unsurprisingly larger body size and high social capital are qualities often associated with leaders in small-scale societies [@glowacki_leadership_2015]. With greater population density and wealth inequality, these risks are expected to concomitantly increase in magnitude.

In managing collective action and enacting group-level policies, some behaviors of leaders will upset some followers and leader-follower interactions can be hostile.

Conceptualizing leadership as a possible solution to problems associated with collective goods [@olson_logic_1965], where group members pool resources and share costs of public goods to receive a net benefit from a cooperative endeavor, leader incur costs associated with managing contributions, coordinating behaviors, and broadcasting information to maintain coordination and cooperation. In this framework, leaders and followers voluntary participate in a cooperative enterprise with the expectation that benefits will outweigh the imposed costs. @rich_dynamics_1980 adopts this framework of collective costs and benefits to understand leadership and group dynamics in urban neighborhoods and neighborhood organizations, which can be viewed as self-interested groups which face group-wide problems and necessarily share public benefits. The costs associated with leadership in neighborhood organizations follow those associated with other collective enterprises (e.g., investments of time and energy in management and maintenance of group coordination). @rich_dynamics_1980 conceptualizes four categories of the benefits associated with neighborhood organization. Leaders of neighborhood organizations (1) may receive benefits of the public goods produces from collective actions (as other non-leaders would as well); (2), may receive selective benefits from surpluses produced through collective actions; (3) may receive incidental benefits from their leadership role; and (4) leaders of neighborhood organizations may enjoy their service in the position. @rich_dynamics_1980 interviewed 54 officers from 11 metropolitan neighborhood organizations in the Midwest US. Results from extensive interviews suggest personal fulfillment and enjoyment are the main motivational drivers of participation in leadership roles of neighborhood organizations, suggesting the process of leadership for individuals is distinct from theoretical models rooted in economics. Therefore, a primary benefit of leadership, in some contexts, is likely to be intrinsic personal fulfillment and the opportunity to execute personal values of community service.

@@

The type of group to which a leader text referred was coded into seven categories. Figure \@ref(fig:groupsex) indicates the proportion of text records that were categorized into each of these groups, broken out by mode of subsistence. We rely on a modified classification of subsistence from the HRAF, collapsing “hunter-gatherers” and “primarily hunter-gatherers” into a single category, “hunter-gatherers,” and by collapsing “agro-pastoralists” and “pastoralists” into a single “pastoralists” category. We renamed the eHRAF’s “Other” subsistence category, which is applied to societies with various combinations of hunting, fishing, gathering, pastoralism, and horticulture or intensive agriculture, as “mixed.” This produced a subsistence classification with five groups: Hunter-gatherers, Pastoralists, Horticulturalists, Agriculturalists, and Mixed.

@@

@@

The most common *functions* leaders serve, which are documented in over 70% of cultures, are resolving conflicts, providing various social functions, providing counsel or direction in follower's affairs, and organizing various cooperative activities. Leaders also commonly serve as a representative in inter-group interactions, manage the distribution of resources, and provide context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punish other group members and manage economic systems. Notable infrequent functions identified here include controlling group movement and migrations.

The most common *qualities* of leaders are being of high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of the cultures sampled here leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, and maintained strong authority over followers. Notable infrequent qualities of leaders in these data include charisma and fairness.

@@

Results of the current study represent the broadest and most comprehensive view of the qualities of leaders, (e.g., what types of people occupy leadership roles?), the functions of leaders, (e.g., what do leaders do?), and the costs and benefits of leadership for both leaders and followers, across a diverse and culturally representative sample of the ethnographic record. These results suggest universal dimensions of human leadership while also suggesting important variation by group type, subsistence strategy, and sex of the leader.

## The universality of human leadership systems

Across a broad range of cultures and social contexs leaders are nearly universally high status males described as knoweldage or intelligent who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizign cooperation among followers, and providing counsel or input in followers' affairs.

@@

This study is based on the ethnographic record. Our methods are designed to capture a broad range of ethnographic descriptions from a culturally representative, systematic sample. Nonetheless, these results are conditional upon the content ethnographers chose to discuss. The ethnographic record is certainly an imperfect source of information on human cultural diversity and hence a limitation of the current results. We, however, view the existence of the ethnographic record and the availability and functionality of the eHRAF as an enormous benefit for comparative research, given many of the populations represented in these data no longer exist or live as they did during the time of much ethnographic description. Within these ethnographic descriptions there is a general male bias, which could reflect a male bias in community leadership or the more systemic male bias in the ethnographic record. These data are also biased towards cultural models (rather than descriptions of observed behavior) and community leadership; descriptions of kin group and residential group leadership are less frequent.

>>>>

We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study, before presenting results and interpretations. Through novel empirical and exploratory analyses encompassing the full range of cultural diversity and social contexts, we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

@@

Leaders emerge in all human societies. That is, within every population there are particular contexts in which one individual maintains a disproportionate level of influence in the group and functions to facilitate group goals and provides services to the group. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @lewis_leaders_1974; @von_rueden_leadership_2015].

Despite the universality of the phenomenon of leadership, the nature of leadership across populations and across social contexts is highly diverse and the emergence of leaders and the functions they serve are closely tied to group structure. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders tends to emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, specialized skills or knowledge, and social support to achieve or maintain community influence [@kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of collective actions conflict resolution [@glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies, at the community level, representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of leaders. Within communities, however, leaders may emerge whenever individuals for groups. The role of leaders in ritual contexts, such as in initiations, and in providing social services, including facilitating marriages and organizing social activities, are also expected to be common functions leaders serve.

Increased sedentarization and defensibility of resources is often associated with both social stratification and institutionalized managerial roles including stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. In the wake of economic intensification, the functions of leadership becomes increasingly associated with the managing economic systems and distribution of resources. Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on dispersed, less defensible resources [@lopez_evolutionary_2017].

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service and nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions associated with their role. The structure of such organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

Scholars in managerial studies have long discussed the behaviors of leaders and managers have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [@bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leadership functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, creating structures for accomplishing group goals, defending institutional integrity, and maintain internal consistency.

@@

NEED CONCLUSION PARAGRAPH/STATEMENT.

@@

Across culturally and economically diverse populations there are some broad commonalities in the qualities of community leaders. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have differential influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have increased influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

<!-- Other attributes of charismatic leadership found to be culturally contingent included ambitiousness, self-effacing, enthusiasm, and risk-taking [@den_hartog_culture_1999].  -->

@@

In the context of institutionalized social stratification, such as chiefdoms and many agricultural populations, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites, who maintain influence over community decisions are endowed with a high degree of social status rooted in ideologies and associated with their roles or kin group [@stanish_evolution_2010]. In such contexts community leaders often lack strong coercive authority, but maintain substantial influence through ideological systems and ritual or religious prowess inaccessible to non-elites [@stanish_evolution_2010]. The association of formal community leadership and supernatural abilities is likely an important precursor of institutionalized and coercive authority.

Within state-level societies including non-industrial kingdoms leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social and political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].

Across disciplines many theoretical models and empirical findings have implicated high intelligence as an important quality positively associated with effective leadership [@johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009]. Recent cross-cultural evidence and from a small-scale society implicates strong relationships between perceptions of intelligence and leadership among males [@garfield_evolutionary_2019; @garfield_investigating_2019], when considered in the context of broad interdisciplinary results suggests high intelligence and specialized knowledge are universal traits of political and community leaders. Recent evidence from within managerial studies, however, suggests a curvilinear effect in that exceptionally intelligence individuals at high end of the distribution may not be the most effective leaders in contemporary business settings [@antonakis_can_2017].

Within managerial studies scholars often suggest the cognitive demands of leadership escalate with increasing organizational complexity [@zaccaro_organizational_2002]. More complex organizations require leaders equipped to process complex information, understand distinct departmentalized bodies, and resolve conflicting points of view, as well as be equipped to socially manage diverse individuals [@bass_handbook_1990; @schneider_organizations_2006; @antonakis_does_2009].

<!-- COULD GET INTO LEADER VS MANAGER DISTINCTION -->

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners [@garfield_evolutionary_2019; @pietraszewski_evolution_2019; @hagen_leadership_2019].

<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formalized institutionalized roles and informal ephemeral roles, are also often if not universally associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional, but leadership roles through their association with increased social status are widely associated with numerous social, material, and reproductive benefits [@kantner_identifying_2010].

<!-- BENEFITS: SOCIAL BENEFITS, SOCIAL NETWORK -->
Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many industries and organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].

@@

<!-- BENFITS: MATING -->
Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel metaanalysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIST -->
@von_rueden_leadership_2015 review a large body of theory and ethnography and outline four possible mechanisms by which leaders in small-scale societies might offset the potential costs associated with various leadership roles; leaders may, (1) facilitate collective actions that yield net benefits to themselves or their close kin; (2) claim a fee for their services, such a tax or greater share of returns; (3) receive reciprocal exchange in other currencies from followers who have benefited from their services; or (4) may gain other social or reproductive benefits by signaling their high quality as a mate and social partner.

<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models of leadership generally assume the benefits of leadership outweigh the costs, for both leaders and followers. However, the unique costs associated with various leadership positions have received less attention, relative to the benefits of leadership, across disciplines. Group responsibilities necessarily impose opportunity costs on individual priorities and agendas, given leadership responsibilities require investments of time, energy, and resources that could be invested elsewhere.

<!-- COSTS: INCREASED SOCIAL OR PHYSICAL RISK FROM MEDIATION - SSS -->
Mediating conflicts between individuals can be physically and socially costly [@wiessner_power_2010]. Devising and proposing compromises and optimal outcomes can be cognitively challenging and the solutions leaders propose are unlikely to be equally satisfactory for everyone involved. Therefore, leaders stand to become the targets of animosity and social ridicule. In the absence of state-based systems and institutionalized inequalities, all group members are on relative equal footing and equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including overt physical aggression [@boehm_purposive_2008]. Physically formidable individuals with strong social networks are likely to be able o bear these costs more-so than their counterparts and unsurprisingly larger body size and high social capital are qualities often associated with leaders in small-scale societies [@glowacki_leadership_2015]. With greater population density and wealth inequality, these risks are expected to concomitantly increase in magnitude. In managing collective action and enacting group-level policies, some behaviors of leaders will upset some followers and leader-follower interactions can be hostile.

<!-- COSTS: INCREASED SOCIAL OR PHYSICAL RISK FROM MEDIATION - LSS -->
Conceptualizing leadership as a possible solution to problems associated with collective goods [@olson_logic_1965], where group members pool resources and share costs of public goods to receive a net benefit from a cooperative endeavor, leader incur costs associated with managing contributions, coordinating behaviors, and broadcasting information to maintain coordination and cooperation. In this framework, leaders and followers voluntary participate in a cooperative enterprise with the expectation that benefits will outweigh the imposed costs. @rich_dynamics_1980 adopts this framework of collective costs and benefits to understand leadership and group dynamics in urban neighborhoods and neighborhood organizations, which can be viewed as self-interested groups which face group-wide problems and necessarily share public benefits. The costs associated with leadership in neighborhood organizations follow those associated with other collective enterprises (e.g., investments of time and energy in management and maintenance of group coordination). @rich_dynamics_1980 conceptualizes four categories of the benefits associated with neighborhood organization. Leaders of neighborhood organizations (1) may receive benefits of the public goods produces from collective actions (as other non-leaders would as well); (2), may receive selective benefits from surpluses produced through collective actions; (3) may receive incidental benefits from their leadership role; and (4) leaders of neighborhood organizations may enjoy their service in the position. @rich_dynamics_1980 interviewed 54 officers from 11 metropolitan neighborhood organizations in the Midwest US. Results from extensive interviews suggest personal fulfillment and enjoyment are the main motivational drivers of participation in leadership roles of neighborhood organizations, suggesting the process of leadership for individuals is distinct from theoretical models rooted in economics. Therefore, a primary benefit of leadership, in some contexts, is likely to be intrinsic personal fulfillment and the opportunity to execute personal values of community service.

@@

The type of group to which a leader text referred was coded into seven researcher-coded categories: residential groups, kin groups, economic groups, political groups at the community level, political group beyond the community level, military groups, and religious groups. Figure \@ref(fig:groupsex) indicates the proportion of text records that were categorized into each of these groups, broken out by mode of subsistence. We rely on a modified classification of subsistence from the HRAF, collapsing “hunter-gatherers” and “primarily hunter-gatherers” into a single category, “hunter-gatherers,” and by collapsing “agro-pastoralists” and “pastoralists” into a single “pastoralists” category. We renamed the eHRAF’s “Other” subsistence category, which is applied to societies with various combinations of hunting, fishing, gathering, pastoralism, and horticulture or intensive agriculture, as “mixed.” This produced a subsistence classification with five groups: Hunter-gatherers, Pastoralists, Horticulturalists, Agriculturalists, and Mixed.

@@

As Figure \@ref(fig:groupsex) reveals, the distribution of text records by group type varies across subsistence type. Across all subsistence groups, the most frequent group type is political groups at the community level, with the exception of agriculturalitsts in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. Leadership within residential groups is most frequently associated hunter gatherers (`r hg_residential_pct`% of the total sample of hunter gatherer text records) and leadership within kin groups is most frequently associated with hunter gatherers (`r hg_kin_pct`% of the total hunter gatherers sample) and horticulturalists (`r hort_kin_pct`% of the total horticulturalists sample).

<!-- SHOULD WE PUT THE ENTIRE TABLE IN SI? -->

@@

The most common *functions* leaders serve, which are documented in over 70% of cultures, are resolving conflicts, providing various social functions, providing counsel or direction in follower's affairs, and organizing various cooperative activities. Leaders also commonly serve as a representatives in inter-group interactions, manage the distribution of resources, and provide context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punish other group members and manage economic systems. Notable infrequent functions include controlling group movement and migrations.

The most common *qualities* of leaders are being of high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of the cultures sampled here leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, or maintained strong authority over followers. Notable infrequent qualities of leaders in these data include charisma and fairness.

@@

Results of the current study represent the broadest and most comprehensive view of the qualities of leaders, (e.g., the types of people that occupy leadership roles), the functions of leaders, (e.g., what do leaders do), and the costs and benefits of leadership for both leaders and followers, across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence strategy, and sex of the leader.

## The universality of human leadership

Across a broad range of cultures and social contexts leaders are nearly universally high status males described as knowledgable or intelligent who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, and providing counsel or input in followers' affairs. See Figure \@ref(fig:varsupport).

At all subsistence levels leaders emerge across a broad range of group types, including at lower-order social units such as within residential groups and kin groups; within specialized groups such as religous, economic, or military groups; and in political contexts at the community level and beyond. Political leadership at the commmunity level is widespread across all subsistence categories. Economic, military, and religious group leadership, although relatively less frequent, are fairly consistently represented across subsistence levels. See Figure \@ref(fig:groupsex).

## The cultural dependancy of human leadership

The distribiution of text records by group type and subsistence suggests leaders emerge within similar social domains among populations within varied subsistance strategies, but at variable frequencies. Hunter-gatherers and pastoralists, populations more likely to expeirence high residential mobility exhibited greater relative frequencies of leaderhsip wtihin residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community mores o than hunter-gatherers and on par with horticulturalists. Both pastoralists and horticulturalists are known for elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaboratored political leadership systems.

Figure \@ref(fig:groupsex) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more instituionalized and economic systems transcend the household, leadership within households and residental groups errodes. It is interesting, however, that even among the agricultural populations that are charachterized by leadership systems beyong the community level, leadership within residential and family group persists.

<!-- MAY WANT TO RETHINK INTERPRETATIONS, LOTS OF STORIES HERE -->

## The features of human leadership

@@

This study is based on the ethnographic record. Our methods are designed to capture a broad range of ethnographic descriptions from a culturally representative, systematic sample. Nonetheless, these results are conditional upon the content ethnographers chose to discuss. The ethnographic record is certainly an imperfect source of information on human cultural diversity and hence a limitation of the current results. We, however, view the existence of the ethnographic record and the availability and functionality of the eHRAF as an enormous benefit for comparative research, given many of the populations represented in these data no longer exist or live as they did during the time of much ethnographic description. Within these ethnographic descriptions there is a general male bias, which could reflect a male bias in community leadership or the more systemic male bias in the ethnographic record. These data are also biased towards cultural models (rather than descriptions of observed behavior) and community leadership; descriptions of kin group and residential group leadership are less frequent. The high proportion of text records discussing political group leaders could be attributable to our broad search strategy or the content of the ethnographic record of "leadership" more generally.
